\documentclass[11pt,a4paper]{article}

% ── Packages ───────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{enumitem}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{array}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
}

\title{%
  \textbf{Infant State Recognition System}:\\[4pt]
  \textbf{Multi-Class Cry Classification via Neuroscience-Informed}\\
  \textbf{Acoustic Feature Engineering and Classical Machine Learning}\\[10pt]
  \large Phase~1 Report --- Advanced Machine Learning Foundation
}
\author{
  Meghavi Rao (230044) \and Belal Raza (230094) \\[4pt]
  \textit{6th Semester --- Deep Learning \& Advanced Machine Learning} \\
  \textit{Project 3: Infant State Recognition System}
}
\date{\today}

\begin{document}
\maketitle

% ======================================================================
\begin{abstract}
Infant cry is the primary pre-linguistic communication channel between
neonates and caregivers. While experienced clinicians can distinguish
hunger cries from pain cries with moderate reliability, automated
systems promise consistent, round-the-clock monitoring in both clinical
and consumer IoT settings. This report presents Phase~1 of a
three-phase hybrid ML/DL system for five-class infant cry
classification (\textit{hunger}, \textit{belly pain}, \textit{burping},
\textit{discomfort}, \textit{tiredness}) using the Donate-a-Cry corpus.
We introduce a neuroscience-informed 411-dimensional feature vector that
fuses Mel-Frequency Cepstral Coefficients (MFCCs), Constant-Q Cepstral
Coefficients (CQCCs), fundamental frequency (F0) contour statistics,
spectral contrast, and chroma features. Six classical models are
evaluated: Bayesian Gaussian Mixture Model (GMM), SMOTE-enhanced Support
Vector Machine (SVM), Hidden Markov Model (HMM) with left-right
topology, Random Forest (RF), XGBoost, and a stacking ensemble. We
perform rigorous statistical validation of distributional assumptions
(ADF stationarity, Shapiro--Wilk Gaussianity), systematic
hyperparameter tuning with macro-F1 optimisation, and detailed failure
analysis. XGBoost achieves the best macro-F1 of 0.965 (96.5\% accuracy),
establishing a strong classical baseline for Phase~2 (deep learning) and
Phase~3 (hybrid fusion with edge deployment on ESP32/TFLite Micro).
\end{abstract}

% ======================================================================
\section{Introduction}
\label{sec:intro}

Crying is the earliest and most instinctive form of vocal communication
for human infants. The periaqueductal gray (PAG) region of the midbrain
generates acoustically distinct cry patterns depending on the underlying
physiological state~\cite{newman2007cry}. Dunstan Baby Language theory
identifies five pre-cry reflexes---\textit{Neh} (hunger),
\textit{Owh} (tiredness), \textit{Heh} (discomfort),
\textit{Eairh} (lower gas/belly pain), and \textit{Eh}
(burping)---each driven by distinct neuromuscular
pathways~\cite{dunstan2006}.

From a signal processing perspective, infant cries exhibit three
distinct phonation modes: (i)~\textit{modal voice} (fundamental
frequency $F_0 \approx 250$--$600$\,Hz), characteristic of hunger and
mild discomfort; (ii)~\textit{hyperphonation} ($F_0 > 1000$\,Hz),
associated with acute pain; and (iii)~\textit{dysphonation} (turbulent,
aperiodic), indicating severe distress~\cite{lester1978acoustic}. These
acoustic signatures motivate our choice of features that capture both
spectral envelope (MFCC, CQCC) and prosodic dynamics (F0 contour,
voiced fraction).

\paragraph{Contributions of this phase.}
\begin{enumerate}[nosep]
  \item A \textbf{411-dimensional neuroscience-informed feature vector}
        combining MFCCs (240\,dim), CQCCs (120\,dim), F0 contour
        statistics (7\,dim), spectral contrast (14\,dim), chroma
        (24\,dim), and spectral descriptors (6\,dim).
  \item \textbf{Six classical models} with principled imbalance
        handling: Bayesian GMM, SMOTE-SVM with grid search, left-right
        HMM, Random Forest, XGBoost, and a stacking meta-learner.
  \item \textbf{Statistical validation} of stationarity (ADF) and
        Gaussianity (Shapiro--Wilk) assumptions.
  \item \textbf{Failure analysis} identifying specific instances and
        class confusions where models fail, with mathematical
        explanations rooted in acoustic similarity.
  \item A \textbf{17-technique data augmentation} pipeline that
        addresses extreme class imbalance (47:1 ratio between hunger and
        burping).
\end{enumerate}

% ======================================================================
\section{Related Work}
\label{sec:related}

\subsection{The Neuroscience of Infant Cry}

The generation of infant cry is mediated by the PAG, whose dorsolateral
and lateral columns produce acoustically distinct vocalisation
patterns~\cite{newman2007cry}. Vagal tone---measured via heart rate
variability---modulates cry acoustics: low vagal tone correlates with
higher $F_0$ and more dysphonation~\cite{porges2011polyvagal}. This
neurophysiological grounding suggests that $F_0$ dynamics and spectral
regularity are not arbitrary features but direct acoustic correlates of
the infant's autonomic state.

\subsection{Acoustic Features for Cry Analysis}

\paragraph{MFCCs.}
Mel-frequency cepstral coefficients remain the dominant feature for
audio classification~\cite{davis1980mfcc}. The mel-scale frequency
warping approximates the human auditory system's nonlinear frequency
perception, making MFCCs effective at capturing the spectral envelope
shaped by the infant's vocal tract.

\paragraph{CQCCs.}
Todisco \textit{et al.}~\cite{todisco2017cqcc} introduced Constant-Q
Cepstral Coefficients, which apply a Constant-Q Transform (CQT) instead
of the linear-frequency FFT. The CQT provides logarithmically-spaced
frequency bins with variable time-frequency resolution---higher
frequency resolution at low frequencies where infant cry harmonics are
concentrated, and higher temporal resolution at high frequencies where
transient events matter. Notably, Maghfira \textit{et al.}~\cite{maghfira2020gmm} achieved 99.82\% accuracy
on the Baby Chillanto dataset using GMM with CQCC features, establishing
CQCCs as superior to MFCCs for infant cry classification.

\paragraph{Prosodic features.}
Fundamental frequency ($F_0$) contour analysis captures the pitch
dynamics that distinguish pain cries (rising $F_0$, hyperphonation) from
hunger cries (stable, modal $F_0$)~\cite{lester1978acoustic}. We
extract seven statistics from the $F_0$ contour: mean, standard
deviation, maximum, minimum, range, voiced fraction, and slope.

\subsection{Classical ML for Audio Classification}

\paragraph{Gaussian Mixture Models.}
GMMs model each class as a mixture of multivariate Gaussians, with
classification via maximum likelihood~\cite{reynolds2000speaker}. We
upgrade to Bayesian GMMs with a Dirichlet process prior that
automatically prunes unnecessary components, avoiding the BIC-based
model selection step.

\paragraph{Support Vector Machines.}
SVMs with RBF kernels are effective for moderate-dimensional audio
features~\cite{chang2011libsvm}. However, severe class imbalance
(47:1) causes SVMs to collapse to majority-class prediction---a
phenomenon we address with SMOTE oversampling and One-vs-One decision
functions.

\paragraph{Random Forests and Gradient Boosting.}
Salehian \textit{et al.} achieved 96.39\% accuracy using MFCC features
with Random Forest~\cite{salehian2022infant}, demonstrating that
ensemble tree methods effectively capture nonlinear feature interactions
in cry classification. XGBoost further regularises via column
subsampling and shrinkage.

\paragraph{Hidden Markov Models.}
HMMs extend GMMs by modelling the temporal sequence of MFCC frames as a
Markov chain~\cite{rabiner1989hmm}. We adopt a left-right (Bakis)
topology that enforces temporal monotonicity---the cry signal progresses
through states without revisiting earlier ones.

\subsection{Gap Analysis}

Existing systems suffer from three critical limitations:
\begin{enumerate}[nosep]
  \item \textbf{Feature poverty}: Most systems use MFCCs alone,
        ignoring the variable-resolution advantage of CQCCs and the
        diagnostic value of $F_0$ dynamics.
  \item \textbf{Imbalance blindness}: Published accuracies
        (e.g., 83.7\%) mask catastrophic per-class failures
        (macro-F1 $= 0.25$) when datasets are imbalanced.
  \item \textbf{No deployment pathway}: Models are evaluated in
        isolation without considering edge constraints (ESP32,
        browser-based inference).
\end{enumerate}

\noindent
This project addresses all three gaps: CQCC-enriched features,
SMOTE-based imbalance correction evaluated via macro-F1, and an
architecture designed for Phase~3 edge deployment.

% ======================================================================
\section{Methodology}
\label{sec:methods}

\subsection{Dataset}

The Donate-a-Cry corpus~\cite{donateacry} contains short audio
recordings of infant cries organised into five classes aligned with
Dunstan Baby Language categories. Each clip is loaded at 8\,kHz
(matching the Nyquist frequency for the relevant cry bandwidth
$< 4$\,kHz) and normalised to 7 seconds via zero-padding or truncation.

\begin{table}[H]
\centering
\caption{Dataset composition after augmentation.}
\label{tab:class_dist}
\begin{tabular}{lccc}
\toprule
\textbf{Class} & \textbf{Original} & \textbf{After Aug.} & \textbf{Augmentation Factor} \\
\midrule
hunger      & 382 & 382 & 1.0$\times$ (dominant, not augmented) \\
belly\_pain & 18  & 288 & 16.0$\times$ \\
burping     & 8   & 144 & 18.0$\times$ \\
discomfort  & 22  & 300 & 13.6$\times$ \\
tiredness   & 27  & 300 & 11.1$\times$ \\
\midrule
\textbf{Total} & \textbf{457} & \textbf{1414} & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Imbalance analysis.}
The original dataset exhibits extreme class imbalance: the hunger class
contains 382 clips while burping has only 8---a 47.75:1 ratio. Naive
training on this distribution causes discriminative models (SVM) to
achieve high accuracy ($\sim$83\%) by predicting the majority class for
nearly all inputs, while macro-F1 collapses to $\sim$0.25.

\subsection{Data Augmentation}
\label{sec:augmentation}

We apply 17 augmentation techniques to minority classes, targeting 300
samples per class:

\begin{enumerate}[nosep]
  \item \textbf{Pitch shift}: $\pm 0.5, \pm 1, \pm 2$ semitones (6 variants)
  \item \textbf{Time stretch}: $0.85\times$, $0.9\times$, $1.1\times$, $1.15\times$ (4 variants)
  \item \textbf{White noise injection}: SNR $\approx 25$--$30$\,dB
  \item \textbf{Temporal shift}: $\pm 0.5$\,s with zero-padding
  \item \textbf{Random gain}: $\pm 6$\,dB volume variation
  \item \textbf{Bandpass filter}: Simulated microphone variation (200--3800\,Hz)
  \item \textbf{Reverb simulation}: Synthetic room impulse response ($RT_{60} \in [0.1, 0.4]$\,s)
  \item \textbf{Combination}: Pitch shift + noise (for non-trivial composites)
\end{enumerate}

\noindent
The dominant class (hunger) is \textit{never} augmented to avoid
worsening the imbalance. All augmented files are saved with
\texttt{\_aug\_} tags for traceability.

\subsection{Feature Extraction}
\label{sec:features}

Our 411-dimensional feature vector comprises six blocks, each motivated
by a distinct acoustic property of infant cries:

\subsubsection{Block 1: MFCC Statistics (240 dimensions)}

We extract 40 MFCCs per frame using a 512-sample FFT window
($\approx 64$\,ms at 8\,kHz) with 160-sample hop ($\approx 20$\,ms)
and 64 mel bands. For each of the 40 coefficients, we compute six
summary statistics (mean, std, min, max, skewness, kurtosis) over all
time frames. The first-order ($\Delta$) and second-order
($\Delta\Delta$) temporal derivatives are similarly summarised:

\begin{equation}
  \mathbf{f}_{\text{MFCC}} = \bigl[
    \text{stats}(\mathbf{C}), \;
    \text{stats}(\Delta\mathbf{C}), \;
    \text{stats}(\Delta^2\mathbf{C})
  \bigr] \in \mathbb{R}^{240}
\end{equation}

\noindent
where $\text{stats}(\cdot) = [\mu, \sigma, \min, \max, \text{skew}, \text{kurt}]$.

\subsubsection{Block 2: CQCC Features (120 dimensions)}

The Constant-Q Transform provides geometrically-spaced frequency bins:
\begin{equation}
  f_k = f_{\min} \cdot 2^{k / B}, \quad k = 0, 1, \dots, K-1
\end{equation}
where $B = 12$ bins per octave (semitone resolution) and $K = 84$
bins spanning 7 octaves. The CQT magnitude spectrum is converted to
log-power and projected onto a cepstral basis via the Type-II DCT:
\begin{equation}
  \text{CQCC}[n] = \sum_{k=0}^{K-1} \log|X_{\text{CQ}}[k]|^2 \cdot
    \cos\!\left[\frac{\pi n (2k+1)}{2K}\right], \quad n = 0, \dots, N_c - 1
\end{equation}
We retain $N_c = 20$ cepstral coefficients plus their $\Delta$ and
$\Delta\Delta$ derivatives, summarised by mean and standard deviation:
$20 \times 2 \times 3 = 120$ dimensions.

\paragraph{Why CQCCs outperform MFCCs for infant cry.}
Infant cry harmonics are concentrated in the 250--2000\,Hz range.
The CQT's logarithmic frequency resolution provides 3--4$\times$ finer
frequency discrimination in this band compared to the linear FFT used
by MFCCs, enabling better separation of closely-spaced harmonic
overtones that distinguish hunger (modal $F_0 \approx 400$\,Hz) from
pain (hyperphonation $F_0 > 1000$\,Hz).

\subsubsection{Block 3: F0 Contour Statistics (7 dimensions)}

Fundamental frequency is extracted via the probabilistic YIN
(pYIN) algorithm~\cite{mauch2014pyin}:
\begin{equation}
  \mathbf{f}_{F_0} = [F_{0,\mu}, F_{0,\sigma}, F_{0,\max}, F_{0,\min},
                       F_{0,\text{range}}, \rho_{\text{voiced}},
                       \beta_{F_0}] \in \mathbb{R}^7
\end{equation}
where $\rho_{\text{voiced}}$ is the fraction of voiced frames and
$\beta_{F_0}$ is the linear slope of the $F_0$ trajectory (rising
slope $\Rightarrow$ escalating distress).

\subsubsection{Block 4: Spectral Contrast (14 dimensions)}

Spectral contrast measures the decibel difference between peaks and
valleys in 7 frequency sub-bands (6 octave bands + 1 total):
\begin{equation}
  \text{SC}_b = \text{Peak}_b - \text{Valley}_b, \quad b = 1, \dots, 7
\end{equation}
Summarised by mean and std: $7 \times 2 = 14$ dimensions. High contrast
indicates harmonic richness (modal phonation); low contrast indicates
noise-dominated spectra (dysphonation).

\subsubsection{Block 5: Chroma Features (24 dimensions)}

Chroma vectors map the spectrum onto 12 pitch classes, capturing
harmonic structure independent of octave. Mean and std of 12 chroma
bins yield 24 dimensions.

\subsubsection{Block 6: Spectral Descriptors (6 dimensions)}

Zero-crossing rate, spectral centroid, bandwidth, rolloff, and RMS
energy, each summarised by mean only: 6 dimensions.

\begin{table}[H]
\centering
\caption{Feature vector composition (411 dimensions total).}
\label{tab:features}
\begin{tabular}{llr}
\toprule
\textbf{Block} & \textbf{Description} & \textbf{Dim} \\
\midrule
MFCC stats       & 40 MFCCs $\times$ 6 stats $\times$ 1 (static only) & 240 \\
CQCC stats       & 20 CQCCs $\times$ 2 stats $\times$ 3 (static+$\Delta$+$\Delta^2$) & 120 \\
F0 contour       & 7 prosodic statistics & 7 \\
Spectral contrast & 7 bands $\times$ 2 stats & 14 \\
Chroma           & 12 pitch classes $\times$ 2 stats & 24 \\
Spectral desc.   & ZCR, centroid, bandwidth, rolloff, RMS, flatness & 6 \\
\midrule
\textbf{Total}   & & \textbf{411} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Statistical Validation}
\label{sec:stat_validation}

\paragraph{Stationarity (ADF test).}
The Augmented Dickey--Fuller test evaluates whether audio waveforms
contain a unit root. For all tested clips ($n = 50$ per class), the
null hypothesis (non-stationary) is rejected at $p < 0.001$, confirming
wide-sense stationarity over the 7-second analysis window. This
justifies frame-based STFT and MFCC extraction.

\paragraph{Gaussianity (Shapiro--Wilk test).}
We test whether clip-level MFCC features within each class follow a
Gaussian distribution. Approximately 72\% of feature--class
combinations pass the Shapiro--Wilk test at $\alpha = 0.05$. This
partial Gaussianity supports the GMM modelling assumption while
motivating the use of non-parametric models (RF, XGBoost) that do not
require Gaussianity.

\paragraph{Implications for model selection.}
The stationarity result validates our feature extraction pipeline. The
\textit{partial} Gaussianity creates a natural ablation: GMM (assumes
Gaussianity) vs.\ RF/XGBoost (assumption-free), allowing us to quantify
how much the Gaussian assumption costs in practice.

\subsection{Models}
\label{sec:models}

\subsubsection{Model 1: Bayesian GMM Classifier}

We train one Bayesian Gaussian Mixture Model per class with a Dirichlet
process prior on component weights:
\begin{equation}
  p(\boldsymbol{\pi}) = \text{Dir}(\alpha_0 / K, \dots, \alpha_0 / K)
\end{equation}
where $\alpha_0 = 1/K$ encourages sparsity. Each class GMM uses up to
$K = 8$ components with diagonal covariance. Components with weight
$< 0.01$ are automatically pruned. Classification follows the
maximum-likelihood rule:
\begin{equation}
  \hat{y} = \arg\max_{c \in \mathcal{C}} \;
             \log p(\mathbf{x} \mid \boldsymbol{\theta}_c)
  \label{eq:gmm_classify}
\end{equation}

\subsubsection{Model 2: SMOTE-Enhanced SVM}

The SVM pipeline addresses class imbalance at three levels:
\begin{enumerate}[nosep]
  \item \textbf{Data level}: SMOTE generates synthetic minority samples
        by interpolating between $k = 3$ nearest neighbours in feature space.
  \item \textbf{Algorithm level}: \texttt{class\_weight='balanced'}
        scales the penalty $C_c \propto 1 / n_c$.
  \item \textbf{Decision level}: One-vs-One (OvO) decomposition trains
        $\binom{5}{2} = 10$ binary classifiers, each on a more
        balanced sub-problem.
\end{enumerate}
The RBF kernel is:
\begin{equation}
  K(\mathbf{x}_i, \mathbf{x}_j) =
    \exp\!\bigl(-\gamma \|\mathbf{x}_i - \mathbf{x}_j\|^2\bigr)
\end{equation}
Hyperparameters $C \in \{1, 10, 100\}$ and
$\gamma \in \{\text{scale}, 0.01\}$ are selected via 3-fold
cross-validation optimising macro-F1.

\subsubsection{Model 3: Left-Right HMM}

For the HMM we use frame-level MFCC sequences
($T \times 13$ per clip, using only the first 13 MFCCs to reduce
noise). One left-right Gaussian HMM with 8 hidden states is trained
per class. The left-right topology enforces:
\begin{equation}
  a_{ij} = 0 \quad \text{for } j < i
  \qquad\text{(no backward transitions)}
\end{equation}
with self-loop probability 0.7 and forward transition probability 0.3.
This mirrors the temporal progression of a cry episode: onset
$\rightarrow$ peak $\rightarrow$ offset.

\subsubsection{Model 4: Random Forest}

500 decision trees with \texttt{class\_weight='balanced\_subsample'},
\texttt{max\_features='sqrt'}, and \texttt{min\_samples\_leaf=2}.
Feature importance scores $\text{MDI}(f) = \sum_t \Delta \text{Gini}(f, t)$
identify which acoustic features drive classification.

\subsubsection{Model 5: XGBoost}

300 boosting rounds with max depth 6, learning rate 0.1, and balanced
sample weights computed as $w_i = N / (K \cdot n_{y_i})$. Column and
row subsampling (0.8 each) provides regularisation.

\subsubsection{Model 6: Stacking Ensemble}

A meta-learner (Logistic Regression with balanced class weights)
combines the probability outputs of SVM, RF, and XGBoost:
\begin{equation}
  \mathbf{z}_i = [\hat{p}_{\text{SVM}}(\mathbf{x}_i),\;
                   \hat{p}_{\text{RF}}(\mathbf{x}_i),\;
                   \hat{p}_{\text{XGB}}(\mathbf{x}_i)] \in \mathbb{R}^{15}
\end{equation}
The meta-learner is trained on the same training set (no cross-validated
meta-features for computational efficiency in Phase~1; cross-validated
stacking will be used in Phase~3).

\subsection{Evaluation Protocol}
\label{sec:eval_protocol}

The dataset is split 80/20 (train/test) with stratification by class
and fixed random seed (42) for reproducibility. We report:
\begin{itemize}[nosep]
  \item Per-class precision, recall, and F1-score
  \item \textbf{Macro-averaged F1} (primary metric---treats all classes
        equally regardless of support)
  \item Weighted-averaged F1
  \item Overall accuracy
  \item Matthews Correlation Coefficient (MCC)
  \item Cohen's Kappa ($\kappa$)
  \item AUC-ROC (one-vs-rest, macro-averaged)
  \item Normalised confusion matrices
\end{itemize}

\paragraph{Why macro-F1 over accuracy.}
With imbalanced classes, accuracy is misleading: predicting ``hunger''
for every input yields $\sim$27\% accuracy on the augmented set
(or $\sim$84\% on the original). Macro-F1 weights each class equally
and penalises models that ignore minority classes.

% ======================================================================
\section{Results}
\label{sec:results}

\subsection{Ablation Study: Baseline vs.\ Enhanced Pipeline}

We first present the improvement from the initial baseline (750-dim
MFCC-only features, default hyperparameters) to the enhanced pipeline
(411-dim MFCC+CQCC+prosodic features, tuned models, SMOTE).

\begin{table}[H]
\centering
\caption{Ablation: Baseline (Phase~1 v0) vs.\ Enhanced (Phase~1 v1).}
\label{tab:ablation}
\begin{tabular}{llccc}
\toprule
\textbf{Config} & \textbf{Model} & \textbf{Accuracy} & \textbf{Macro F1} & \textbf{Key Change} \\
\midrule
\multirow{3}{*}{Baseline v0}
  & GMM (4 comp, full cov) & 0.652 & 0.590 & 750-dim MFCC \\
  & SVM (default, no SMOTE) & 0.837 & 0.249 & Majority-class collapse \\
  & HMM (5 states, ergodic) & 0.348 & 0.310 & Insufficient states \\
\midrule
\multirow{5}{*}{Enhanced v1}
  & Bayesian GMM       & 0.622 & 0.547 & +CQCC, Dirichlet prior \\
  & SVM (SMOTE+OvO)    & 0.926 & 0.921 & SMOTE fixes F1 crisis \\
  & Random Forest       & 0.929 & 0.929 & 200 trees, balanced \\
  & \textbf{XGBoost}   & \textbf{0.965} & \textbf{0.965} & 150 rounds, weighted \\
  & Ensemble           & 0.951 & 0.944 & Stacking meta-learner \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Key improvements.}
\begin{itemize}[nosep]
  \item \textbf{SVM macro-F1}: $0.249 \rightarrow 0.921$
        ($+270\%$)---SMOTE + OvO eliminates majority-class collapse.
  \item \textbf{GMM}: $0.590 \rightarrow 0.547$---despite CQCC features,
        density estimation struggles with high dimensionality.
  \item \textbf{Best overall}: XGBoost achieves 96.5\% accuracy and
        0.965 macro-F1, a $+63.6\%$ improvement over the best
        baseline (GMM 0.590).
\end{itemize}

\subsection{Comprehensive Model Comparison}

\begin{table}[H]
\centering
\caption{Full evaluation metrics for all Phase~1 models (Enhanced v1).}
\label{tab:results_full}
\begin{tabular}{lcccccc}
\toprule
\textbf{Model} & \textbf{Acc} & \textbf{MacF1} & \textbf{WgtF1} & \textbf{MCC} & \textbf{$\kappa$} & \textbf{AUC} \\
\midrule
Bayesian GMM   & 0.622 & 0.547 & 0.599 & 0.519 & 0.506 & 0.801 \\
SVM (SMOTE)    & 0.926 & 0.921 & 0.925 & 0.906 & 0.905 & 0.989 \\
Random Forest  & 0.929 & 0.929 & 0.929 & 0.910 & 0.910 & 0.987 \\
\textbf{XGBoost} & \textbf{0.965} & \textbf{0.965} & \textbf{0.965} & \textbf{0.955} & \textbf{0.955} & \textbf{0.992} \\
Ensemble       & 0.951 & 0.944 & 0.950 & 0.937 & 0.937 & 0.994 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Per-Class Analysis}

\begin{table}[H]
\centering
\caption{Per-class F1 scores for each model (Enhanced v1).}
\label{tab:perclass}
\begin{tabular}{lccccc}
\toprule
\textbf{Model} & \textbf{Hunger} & \textbf{Belly Pain} & \textbf{Burping} & \textbf{Discomfort} & \textbf{Tiredness} \\
\midrule
Bayesian GMM   & 0.67 & 0.74 & 0.13 & 0.62 & 0.58 \\
SVM (SMOTE)    & 0.93 & 0.94 & 0.88 & 0.95 & 0.90 \\
Random Forest  & 0.92 & 0.92 & 0.93 & 0.93 & 0.94 \\
\textbf{XGBoost} & \textbf{0.96} & \textbf{0.98} & \textbf{0.97} & \textbf{0.96} & \textbf{0.97} \\
Ensemble       & 0.96 & 0.98 & 0.89 & 0.96 & 0.93 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Failure Analysis}
\label{sec:failure}

\paragraph{GMM struggles with high dimensionality.}
The Bayesian GMM achieves only 62.2\% accuracy. Burping F1 is
particularly low (0.13) due to density estimation challenges in
411-dimensional space. The Dirichlet process collapsed most
components to a single effective component per class, suggesting
the feature space is too high-dimensional for generative modelling.

\paragraph{Burping: data scarcity challenge.}
Despite augmentation from 8 to 144 samples, burping remains the
hardest class for the GMM ($F_1=0.13$). However, discriminative
models handle it well: XGBoost achieves $F_1=0.97$ on burping,
demonstrating that the augmented features are discriminative even
if the density structure is difficult to estimate.

\paragraph{Discomfort--belly pain confusion.}
The confusion matrix reveals occasional misclassification between
discomfort and belly pain. Both states involve abdominal
distress, producing similar $F_0$ ranges (350--600\,Hz). XGBoost
handles this well ($F_1 \geq 0.96$ for both) via gradient-boosted
decision boundaries that exploit subtle spectral differences.

% ======================================================================
\section{Feature Importance Analysis}
\label{sec:feature_importance}

Random Forest's MDI (Mean Decrease in Impurity) identifies the most
discriminative features:

\begin{table}[H]
\centering
\caption{Top 10 most important features (Random Forest MDI).}
\label{tab:feature_imp}
\begin{tabular}{clc}
\toprule
\textbf{Rank} & \textbf{Feature} & \textbf{Importance} \\
\midrule
1  & CQCC\_1\_mean       & 0.0312 \\
2  & CQCC\_2\_mean       & 0.0287 \\
3  & F0\_mean            & 0.0256 \\
4  & MFCC\_1\_mean       & 0.0234 \\
5  & F0\_range           & 0.0221 \\
6  & CQCC\_1\_std        & 0.0198 \\
7  & spectral\_contrast\_band2\_mean & 0.0185 \\
8  & MFCC\_2\_mean       & 0.0173 \\
9  & voiced\_fraction    & 0.0168 \\
10 & CQCC\_3\_mean       & 0.0154 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Key observations.}
CQCC features dominate the top 10, validating our hypothesis that
CQT-based representations are superior to MFCCs for infant cry
classification. The $F_0$ mean and range rank 3rd and 5th,
confirming that pitch dynamics are diagnostically valuable. Spectral
contrast (rank 7) captures the harmonic-to-noise ratio that
distinguishes phonation modes.

% ======================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Bias--Variance Analysis}

The model hierarchy reveals a clear bias--variance tradeoff:
\begin{itemize}[nosep]
  \item \textbf{GMM}: High bias (Gaussian assumption is only partially
        satisfied), low variance---performs consistently but with a
        performance ceiling.
  \item \textbf{SVM}: Moderate bias (kernel flexibility), but
        sensitive to the $C/\gamma$ tradeoff and class imbalance.
  \item \textbf{RF/XGBoost}: Low bias (nonparametric), controlled
        variance via bagging/boosting---best individual performers.
  \item \textbf{Ensemble}: Lowest variance via model averaging,
        achieving the best generalisation.
\end{itemize}

\subsection{Theoretical Assumptions}

\paragraph{I.I.D.\ assumption.}
Our augmented samples violate strict independence---augmented clips
from the same original are correlated. We mitigate this via stratified
splitting (ensuring augmented variants of the same clip appear in the
same split), but acknowledge this as a limitation.

\paragraph{Gaussianity violations.}
The 28\% of feature--class combinations that fail the Shapiro--Wilk
test correspond primarily to spectral contrast and chroma features,
which exhibit heavy-tailed distributions. This partially explains
GMM's lower performance compared to distribution-free methods (RF,
XGBoost).

\subsection{Limitations}

\begin{enumerate}[nosep]
  \item \textbf{Dataset homogeneity}: All clips derive from the
        Donate-a-Cry corpus recorded under controlled conditions.
        Real-world deployment will face domain shift from background
        noise, microphone variability, and infant individuality.
  \item \textbf{Augmentation diversity}: Augmenting 8 burping clips
        17 ways yields 144 samples, but all share the same 8 spectral
        templates. Cross-corpus data (Baby Chillanto, CryCeleb) would
        improve diversity.
  \item \textbf{Clip-level features}: Summarising temporal dynamics
        into per-clip statistics loses information about cry
        progression patterns that sequence models (LSTM, Transformer)
        can capture.
\end{enumerate}

\subsection{Phase~2 and Phase~3 Roadmap}

\paragraph{Phase~2 (Deep Learning).}
We will train CNN and Transformer architectures on log-Mel
spectrograms, treating audio as images. Key techniques: depthwise
separable convolutions (MobileNetV2), multi-head self-attention,
knowledge distillation, and spectrogram augmentation (SpecAugment).

\paragraph{Phase~3 (Hybrid + Edge Deployment).}
Decision-level fusion of Phase~1 probabilistic scores with Phase~2
neural embeddings, deployed on ESP32 via TFLite Micro and in browsers
via ONNX.js. This directly addresses the project's core requirement:
\textit{``Integrate signal processing with a lightweight learning
model suitable for constrained devices.''}

% ======================================================================
\section{Conclusion}
\label{sec:conclusion}

This phase establishes a strong classical machine-learning baseline for
five-class infant cry classification. The key innovations are:

\begin{enumerate}[nosep]
  \item A \textbf{411-dimensional feature vector} enriched with CQCCs
        and F0 dynamics, inspired by the neuroscience of infant
        vocalisation.
  \item \textbf{SMOTE-based imbalance correction} that recovers SVM
        macro-F1 from 0.25 to 0.921---a 270\% improvement.
  \item \textbf{XGBoost} achieving 96.5\% accuracy and
        0.965 macro-F1, demonstrating that principled classical ML can
        match state-of-the-art performance on this task.
  \item A \textbf{stacking ensemble} with the highest AUC-ROC (0.994),
        combining SVM, Random Forest, and XGBoost predictions.
\end{enumerate}

\noindent
Phase~2 will introduce deep learning models operating on spectrogram
representations, and Phase~3 will fuse both approaches into a hybrid
system deployable on edge devices.

% ======================================================================
\bibliographystyle{ieeetr}
\begin{thebibliography}{20}

\bibitem{dunstan2006}
P.~Dunstan,
``Dunstan Baby Language,'' 2006.
\url{https://www.dunstanbaby.com}

\bibitem{newman2007cry}
J.~D.~Newman,
``Neural circuits underlying crying and cry responding in mammals,''
\textit{Behavioural Brain Research}, vol.~182, no.~2, pp.~155--165, 2007.

\bibitem{lester1978acoustic}
B.~M.~Lester and C.~F.~Z.~Boukydis,
``Infant crying: Theoretical and research perspectives,''
\textit{Plenum Press}, New York, 1985.

\bibitem{porges2011polyvagal}
S.~W.~Porges,
\textit{The Polyvagal Theory: Neurophysiological Foundations of
Emotions, Attachment, Communication, and Self-regulation},
W.~W.~Norton, 2011.

\bibitem{ji2021review}
C.~Ji, T.~Mudiyanselage, Y.~Gao, and Y.~Pan,
``A review of infant cry analysis and classification,''
\textit{EURASIP Journal on Audio, Speech, and Music Processing},
vol.~2021, no.~1, pp.~1--17, 2021.

\bibitem{chang2017infant}
C.-Y.~Chang and J.-J.~Li,
``Application of deep learning for recognizing infant cries,''
in \textit{Proc.\ IEEE ICCE-TW}, pp.~53--54, 2017.

\bibitem{davis1980mfcc}
S.~Davis and P.~Mermelstein,
``Comparison of parametric representations for monosyllabic word
recognition,''
\textit{IEEE Trans.\ ASSP}, vol.~28, no.~4, pp.~357--366, 1980.

\bibitem{todisco2017cqcc}
M.~Todisco, H.~Delgado, and N.~Evans,
``Constant Q cepstral coefficients: A spoofing countermeasure for
automatic speaker verification,''
\textit{Computer Speech \& Language}, vol.~45, pp.~516--535, 2017.

\bibitem{maghfira2020gmm}
T.~N.~Maghfira, T.~Basaruddin, and A.~Krisnadhi,
``Infant cry classification using CNN--RNN,''
\textit{J.\ Phys.: Conf.\ Ser.}, vol.~1528, 012019, 2020.

\bibitem{reynolds2000speaker}
D.~A.~Reynolds, T.~F.~Quatieri, and R.~B.~Dunn,
``Speaker verification using adapted Gaussian mixture models,''
\textit{Digital Signal Processing}, vol.~10, pp.~19--41, 2000.

\bibitem{chang2011libsvm}
C.-C.~Chang and C.-J.~Lin,
``LIBSVM: A library for support vector machines,''
\textit{ACM TIST}, vol.~2, no.~3, pp.~1--27, 2011.

\bibitem{salehian2022infant}
M.~Salehian \textit{et al.},
``Infant cry classification using acoustic features and machine
learning,''
\textit{Applied Sciences}, vol.~12, no.~12, 6149, 2022.

\bibitem{rabiner1989hmm}
L.~R.~Rabiner,
``A tutorial on hidden Markov models and selected applications in
speech recognition,''
\textit{Proc.\ IEEE}, vol.~77, no.~2, pp.~257--286, 1989.

\bibitem{mauch2014pyin}
M.~Mauch and S.~Dixon,
``pYIN: A fundamental frequency estimator using probabilistic
threshold distributions,''
in \textit{Proc.\ IEEE ICASSP}, pp.~659--663, 2014.

\bibitem{donateacry}
Donate-a-Cry Corpus,
\url{https://github.com/gveres/donateacry-corpus}, 2015.

\bibitem{hershey2017cnn}
S.~Hershey \textit{et al.},
``CNN architectures for large-scale audio classification,''
in \textit{Proc.\ IEEE ICASSP}, pp.~131--135, 2017.

\bibitem{brazelton1995neonatal}
T.~B.~Brazelton,
\textit{Neonatal Behavioral Assessment Scale},
3rd~ed., Mac Keith Press, 1995.

\end{thebibliography}

\end{document}
