\documentclass[11pt,a4paper]{article}

% ── Packages ───────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{enumitem}
\usepackage{float}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{array}

\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
}

\title{%
  \textbf{Infant State Recognition System}:\\[4pt]
  \textbf{Multi-Class Cry Classification via Neuroscience-Informed}\\
  \textbf{Acoustic Feature Engineering and Classical Machine Learning}\\[10pt]
  \large Phase~1 Report --- Advanced Machine Learning Foundation\\
  \large with Data Leakage Analysis and Correction
}
\author{
  Meghavi Rao (230044) \and Belal Raza (230094) \\[4pt]
  \textit{6th Semester --- Deep Learning \& Advanced Machine Learning} \\
  \textit{Project 3: Infant State Recognition System}
}
\date{\today}

\begin{document}
\maketitle

% ======================================================================
\begin{abstract}
Infant cry is the primary pre-linguistic communication channel between
neonates and caregivers. While experienced clinicians can distinguish
hunger cries from pain cries with moderate reliability, automated
systems promise consistent, round-the-clock monitoring in both clinical
and consumer IoT settings. This report presents Phase~1 of a
three-phase hybrid ML/DL system for five-class infant cry
classification (\textit{hunger}, \textit{belly pain}, \textit{burping},
\textit{discomfort}, \textit{tiredness}) using the Donate-a-Cry corpus.
We introduce a neuroscience-informed 411-dimensional feature vector that
fuses Mel-Frequency Cepstral Coefficients (MFCCs), Constant-Q Cepstral
Coefficients (CQCCs), fundamental frequency (F0) contour statistics,
spectral contrast, and chroma features. Six classical models are
evaluated: Bayesian Gaussian Mixture Model (GMM), SMOTE-enhanced Support
Vector Machine (SVM), Hidden Markov Model (HMM) with left-right
topology, Random Forest (RF), XGBoost, and a stacking ensemble. We
perform rigorous statistical validation of distributional assumptions
(ADF stationarity, Shapiro--Wilk Gaussianity), systematic
hyperparameter tuning with macro-F1 optimisation, and detailed failure
analysis. Critically, we identify and correct a \textbf{data leakage}
issue where augmented audio variants leaked across the train/test split,
inflating initial results (XGBoost: 96.5\% accuracy). After correction
via a split-before-augment protocol, the Ensemble achieves 83.7\%
accuracy and SVM achieves the best macro-F1 of 0.270 on the
leak-free test set of 92 original recordings. This honest evaluation
reveals the true challenge posed by extreme class imbalance (47:1 ratio)
and establishes a rigorous baseline for Phase~2 (deep learning) and
Phase~3 (hybrid fusion with edge deployment).
\end{abstract}

% ======================================================================
\section{Introduction}
\label{sec:intro}

Crying is the earliest and most instinctive form of vocal communication
for human infants. The periaqueductal gray (PAG) region of the midbrain
generates acoustically distinct cry patterns depending on the underlying
physiological state~\cite{newman2007cry}. Dunstan Baby Language theory
identifies five pre-cry reflexes---\textit{Neh} (hunger),
\textit{Owh} (tiredness), \textit{Heh} (discomfort),
\textit{Eairh} (lower gas/belly pain), and \textit{Eh}
(burping)---each driven by distinct neuromuscular
pathways~\cite{dunstan2006}.

From a signal processing perspective, infant cries exhibit three
distinct phonation modes: (i)~\textit{modal voice} (fundamental
frequency $F_0 \approx 250$--$600$\,Hz), characteristic of hunger and
mild discomfort; (ii)~\textit{hyperphonation} ($F_0 > 1000$\,Hz),
associated with acute pain; and (iii)~\textit{dysphonation} (turbulent,
aperiodic), indicating severe distress~\cite{lester1978acoustic}. These
acoustic signatures motivate our choice of features that capture both
spectral envelope (MFCC, CQCC) and prosodic dynamics (F0 contour,
voiced fraction).

\paragraph{Contributions of this phase.}
\begin{enumerate}[nosep]
  \item A \textbf{411-dimensional neuroscience-informed feature vector}
        combining MFCCs (240\,dim), CQCCs (120\,dim), F0 contour
        statistics (7\,dim), spectral contrast (14\,dim), chroma
        (24\,dim), and spectral descriptors (6\,dim).
  \item \textbf{Six classical models} with principled imbalance
        handling: Bayesian GMM, SMOTE-SVM with grid search, left-right
        HMM, Random Forest, XGBoost, and a stacking meta-learner.
  \item \textbf{Statistical validation} of stationarity (ADF) and
        Gaussianity (Shapiro--Wilk) assumptions.
  \item \textbf{Failure analysis} identifying specific instances and
        class confusions where models fail, with mathematical
        explanations rooted in acoustic similarity.
  \item A \textbf{17-technique data augmentation} pipeline that
        addresses extreme class imbalance (47:1 ratio between hunger and
        burping).
\end{enumerate}

% ======================================================================
\section{Related Work}
\label{sec:related}

\subsection{The Neuroscience of Infant Cry}

The generation of infant cry is mediated by the PAG, whose dorsolateral
and lateral columns produce acoustically distinct vocalisation
patterns~\cite{newman2007cry}. Vagal tone---measured via heart rate
variability---modulates cry acoustics: low vagal tone correlates with
higher $F_0$ and more dysphonation~\cite{porges2011polyvagal}. This
neurophysiological grounding suggests that $F_0$ dynamics and spectral
regularity are not arbitrary features but direct acoustic correlates of
the infant's autonomic state.

\subsection{Acoustic Features for Cry Analysis}

\paragraph{MFCCs.}
Mel-frequency cepstral coefficients remain the dominant feature for
audio classification~\cite{davis1980mfcc}. The mel-scale frequency
warping approximates the human auditory system's nonlinear frequency
perception, making MFCCs effective at capturing the spectral envelope
shaped by the infant's vocal tract.

\paragraph{CQCCs.}
Todisco \textit{et al.}~\cite{todisco2017cqcc} introduced Constant-Q
Cepstral Coefficients, which apply a Constant-Q Transform (CQT) instead
of the linear-frequency FFT. The CQT provides logarithmically-spaced
frequency bins with variable time-frequency resolution---higher
frequency resolution at low frequencies where infant cry harmonics are
concentrated, and higher temporal resolution at high frequencies where
transient events matter. Notably, Maghfira \textit{et al.}~\cite{maghfira2020gmm} achieved 99.82\% accuracy
on the Baby Chillanto dataset using GMM with CQCC features, establishing
CQCCs as superior to MFCCs for infant cry classification.

\paragraph{Prosodic features.}
Fundamental frequency ($F_0$) contour analysis captures the pitch
dynamics that distinguish pain cries (rising $F_0$, hyperphonation) from
hunger cries (stable, modal $F_0$)~\cite{lester1978acoustic}. We
extract seven statistics from the $F_0$ contour: mean, standard
deviation, maximum, minimum, range, voiced fraction, and slope.

\subsection{Classical ML for Audio Classification}

\paragraph{Gaussian Mixture Models.}
GMMs model each class as a mixture of multivariate Gaussians, with
classification via maximum likelihood~\cite{reynolds2000speaker}. We
upgrade to Bayesian GMMs with a Dirichlet process prior that
automatically prunes unnecessary components, avoiding the BIC-based
model selection step.

\paragraph{Support Vector Machines.}
SVMs with RBF kernels are effective for moderate-dimensional audio
features~\cite{chang2011libsvm}. However, severe class imbalance
(47:1) causes SVMs to collapse to majority-class prediction---a
phenomenon we address with SMOTE oversampling and One-vs-One decision
functions.

\paragraph{Random Forests and Gradient Boosting.}
Salehian \textit{et al.} achieved 96.39\% accuracy using MFCC features
with Random Forest~\cite{salehian2022infant}, demonstrating that
ensemble tree methods effectively capture nonlinear feature interactions
in cry classification. XGBoost further regularises via column
subsampling and shrinkage.

\paragraph{Hidden Markov Models.}
HMMs extend GMMs by modelling the temporal sequence of MFCC frames as a
Markov chain~\cite{rabiner1989hmm}. We adopt a left-right (Bakis)
topology that enforces temporal monotonicity---the cry signal progresses
through states without revisiting earlier ones.

\subsection{Gap Analysis}

Existing systems suffer from three critical limitations:
\begin{enumerate}[nosep]
  \item \textbf{Feature poverty}: Most systems use MFCCs alone,
        ignoring the variable-resolution advantage of CQCCs and the
        diagnostic value of $F_0$ dynamics.
  \item \textbf{Imbalance blindness}: Published accuracies
        (e.g., 83.7\%) mask catastrophic per-class failures
        (macro-F1 $= 0.25$) when datasets are imbalanced.
  \item \textbf{No deployment pathway}: Models are evaluated in
        isolation without considering edge constraints (ESP32,
        browser-based inference).
\end{enumerate}

\noindent
This project addresses all three gaps: CQCC-enriched features,
SMOTE-based imbalance correction evaluated via macro-F1, and an
architecture designed for Phase~3 edge deployment.

% ======================================================================
\section{Methodology}
\label{sec:methods}

\subsection{Dataset}

The Donate-a-Cry corpus~\cite{donateacry} contains short audio
recordings of infant cries organised into five classes aligned with
Dunstan Baby Language categories. Each clip is loaded at 8\,kHz
(matching the Nyquist frequency for the relevant cry bandwidth
$< 4$\,kHz) and normalised to 7 seconds via zero-padding or truncation.

\begin{table}[H]
\centering
\caption{Dataset composition after augmentation.}
\label{tab:class_dist}
\begin{tabular}{lccc}
\toprule
\textbf{Class} & \textbf{Original} & \textbf{After Aug.} & \textbf{Augmentation Factor} \\
\midrule
hunger      & 382 & 382 & 1.0$\times$ (dominant, not augmented) \\
belly\_pain & 18  & 288 & 16.0$\times$ \\
burping     & 8   & 144 & 18.0$\times$ \\
discomfort  & 22  & 300 & 13.6$\times$ \\
tiredness   & 27  & 300 & 11.1$\times$ \\
\midrule
\textbf{Total} & \textbf{457} & \textbf{1414} & --- \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Imbalance analysis.}
The original dataset exhibits extreme class imbalance: the hunger class
contains 382 clips while burping has only 8---a 47.75:1 ratio. Naive
training on this distribution causes discriminative models (SVM) to
achieve high accuracy ($\sim$83\%) by predicting the majority class for
nearly all inputs, while macro-F1 collapses to $\sim$0.25.

\subsection{Data Augmentation}
\label{sec:augmentation}

We apply 17 augmentation techniques to minority classes, targeting 300
samples per class:

\begin{enumerate}[nosep]
  \item \textbf{Pitch shift}: $\pm 0.5, \pm 1, \pm 2$ semitones (6 variants)
  \item \textbf{Time stretch}: $0.85\times$, $0.9\times$, $1.1\times$, $1.15\times$ (4 variants)
  \item \textbf{White noise injection}: SNR $\approx 25$--$30$\,dB
  \item \textbf{Temporal shift}: $\pm 0.5$\,s with zero-padding
  \item \textbf{Random gain}: $\pm 6$\,dB volume variation
  \item \textbf{Bandpass filter}: Simulated microphone variation (200--3800\,Hz)
  \item \textbf{Reverb simulation}: Synthetic room impulse response ($RT_{60} \in [0.1, 0.4]$\,s)
  \item \textbf{Combination}: Pitch shift + noise (for non-trivial composites)
\end{enumerate}

\noindent
The dominant class (hunger) is \textit{never} augmented to avoid
worsening the imbalance.

\paragraph{Split-before-augment protocol.}
A critical methodological correction was applied after initial
experiments revealed data leakage. In the initial pipeline,
augmented files were written to disk alongside originals, and a
subsequent random train/test split scattered augmented variants of the
same original recording across both sets. This allowed models to be
tested on near-copies of their training data, producing artificially
inflated metrics (e.g., XGBoost: 96.5\% accuracy). The corrected
pipeline enforces:
\begin{enumerate}[nosep]
  \item Load only original recordings (457 files, excluding
        \texttt{\_aug\_} suffixed files).
  \item Split originals into train (365) and test (92) with 80/20
        stratification.
  \item Augment \textit{only} the training set in memory (no disk
        writes).
  \item Extract features separately for train and test.
\end{enumerate}
This ensures the test set contains \textit{exclusively} original,
never-augmented recordings, producing an honest evaluation.

\subsection{Feature Extraction}
\label{sec:features}

Our 411-dimensional feature vector comprises six blocks, each motivated
by a distinct acoustic property of infant cries:

\subsubsection{Block 1: MFCC Statistics (240 dimensions)}

We extract 40 MFCCs per frame using a 512-sample FFT window
($\approx 64$\,ms at 8\,kHz) with 160-sample hop ($\approx 20$\,ms)
and 64 mel bands. For each of the 40 coefficients, we compute six
summary statistics (mean, std, min, max, skewness, kurtosis) over all
time frames. The first-order ($\Delta$) and second-order
($\Delta\Delta$) temporal derivatives are similarly summarised:

\begin{equation}
  \mathbf{f}_{\text{MFCC}} = \bigl[
    \text{stats}(\mathbf{C}), \;
    \text{stats}(\Delta\mathbf{C}), \;
    \text{stats}(\Delta^2\mathbf{C})
  \bigr] \in \mathbb{R}^{240}
\end{equation}

\noindent
where $\text{stats}(\cdot) = [\mu, \sigma, \min, \max, \text{skew}, \text{kurt}]$.

\subsubsection{Block 2: CQCC Features (120 dimensions)}

The Constant-Q Transform provides geometrically-spaced frequency bins:
\begin{equation}
  f_k = f_{\min} \cdot 2^{k / B}, \quad k = 0, 1, \dots, K-1
\end{equation}
where $B = 12$ bins per octave (semitone resolution) and $K = 84$
bins spanning 7 octaves. The CQT magnitude spectrum is converted to
log-power and projected onto a cepstral basis via the Type-II DCT:
\begin{equation}
  \text{CQCC}[n] = \sum_{k=0}^{K-1} \log|X_{\text{CQ}}[k]|^2 \cdot
    \cos\!\left[\frac{\pi n (2k+1)}{2K}\right], \quad n = 0, \dots, N_c - 1
\end{equation}
We retain $N_c = 20$ cepstral coefficients plus their $\Delta$ and
$\Delta\Delta$ derivatives, summarised by mean and standard deviation:
$20 \times 2 \times 3 = 120$ dimensions.

\paragraph{Why CQCCs outperform MFCCs for infant cry.}
Infant cry harmonics are concentrated in the 250--2000\,Hz range.
The CQT's logarithmic frequency resolution provides 3--4$\times$ finer
frequency discrimination in this band compared to the linear FFT used
by MFCCs, enabling better separation of closely-spaced harmonic
overtones that distinguish hunger (modal $F_0 \approx 400$\,Hz) from
pain (hyperphonation $F_0 > 1000$\,Hz).

\subsubsection{Block 3: F0 Contour Statistics (7 dimensions)}

Fundamental frequency is extracted via the probabilistic YIN
(pYIN) algorithm~\cite{mauch2014pyin}:
\begin{equation}
  \mathbf{f}_{F_0} = [F_{0,\mu}, F_{0,\sigma}, F_{0,\max}, F_{0,\min},
                       F_{0,\text{range}}, \rho_{\text{voiced}},
                       \beta_{F_0}] \in \mathbb{R}^7
\end{equation}
where $\rho_{\text{voiced}}$ is the fraction of voiced frames and
$\beta_{F_0}$ is the linear slope of the $F_0$ trajectory (rising
slope $\Rightarrow$ escalating distress).

\subsubsection{Block 4: Spectral Contrast (14 dimensions)}

Spectral contrast measures the decibel difference between peaks and
valleys in 7 frequency sub-bands (6 octave bands + 1 total):
\begin{equation}
  \text{SC}_b = \text{Peak}_b - \text{Valley}_b, \quad b = 1, \dots, 7
\end{equation}
Summarised by mean and std: $7 \times 2 = 14$ dimensions. High contrast
indicates harmonic richness (modal phonation); low contrast indicates
noise-dominated spectra (dysphonation).

\subsubsection{Block 5: Chroma Features (24 dimensions)}

Chroma vectors map the spectrum onto 12 pitch classes, capturing
harmonic structure independent of octave. Mean and std of 12 chroma
bins yield 24 dimensions.

\subsubsection{Block 6: Spectral Descriptors (6 dimensions)}

Zero-crossing rate, spectral centroid, bandwidth, rolloff, and RMS
energy, each summarised by mean only: 6 dimensions.

\begin{table}[H]
\centering
\caption{Feature vector composition (411 dimensions total).}
\label{tab:features}
\begin{tabular}{llr}
\toprule
\textbf{Block} & \textbf{Description} & \textbf{Dim} \\
\midrule
MFCC stats       & 40 MFCCs $\times$ 6 stats $\times$ 1 (static only) & 240 \\
CQCC stats       & 20 CQCCs $\times$ 2 stats $\times$ 3 (static+$\Delta$+$\Delta^2$) & 120 \\
F0 contour       & 7 prosodic statistics & 7 \\
Spectral contrast & 7 bands $\times$ 2 stats & 14 \\
Chroma           & 12 pitch classes $\times$ 2 stats & 24 \\
Spectral desc.   & ZCR, centroid, bandwidth, rolloff, RMS, flatness & 6 \\
\midrule
\textbf{Total}   & & \textbf{411} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Statistical Validation}
\label{sec:stat_validation}

\paragraph{Stationarity (ADF test).}
The Augmented Dickey--Fuller test evaluates whether audio waveforms
contain a unit root. For all tested clips ($n = 50$ per class), the
null hypothesis (non-stationary) is rejected at $p < 0.001$, confirming
wide-sense stationarity over the 7-second analysis window. This
justifies frame-based STFT and MFCC extraction.

\paragraph{Gaussianity (Shapiro--Wilk test).}
We test whether clip-level MFCC features within each class follow a
Gaussian distribution. Approximately 72\% of feature--class
combinations pass the Shapiro--Wilk test at $\alpha = 0.05$. This
partial Gaussianity supports the GMM modelling assumption while
motivating the use of non-parametric models (RF, XGBoost) that do not
require Gaussianity.

\paragraph{Implications for model selection.}
The stationarity result validates our feature extraction pipeline. The
\textit{partial} Gaussianity creates a natural ablation: GMM (assumes
Gaussianity) vs.\ RF/XGBoost (assumption-free), allowing us to quantify
how much the Gaussian assumption costs in practice.

\subsection{Models}
\label{sec:models}

\subsubsection{Model 1: Bayesian GMM Classifier}

We train one Bayesian Gaussian Mixture Model per class with a Dirichlet
process prior on component weights:
\begin{equation}
  p(\boldsymbol{\pi}) = \text{Dir}(\alpha_0 / K, \dots, \alpha_0 / K)
\end{equation}
where $\alpha_0 = 1/K$ encourages sparsity. Each class GMM uses up to
$K = 8$ components with diagonal covariance. Components with weight
$< 0.01$ are automatically pruned. Classification follows the
maximum-likelihood rule:
\begin{equation}
  \hat{y} = \arg\max_{c \in \mathcal{C}} \;
             \log p(\mathbf{x} \mid \boldsymbol{\theta}_c)
  \label{eq:gmm_classify}
\end{equation}

\subsubsection{Model 2: SMOTE-Enhanced SVM}

The SVM pipeline addresses class imbalance at three levels:
\begin{enumerate}[nosep]
  \item \textbf{Data level}: SMOTE generates synthetic minority samples
        by interpolating between $k = 3$ nearest neighbours in feature space.
  \item \textbf{Algorithm level}: \texttt{class\_weight='balanced'}
        scales the penalty $C_c \propto 1 / n_c$.
  \item \textbf{Decision level}: One-vs-One (OvO) decomposition trains
        $\binom{5}{2} = 10$ binary classifiers, each on a more
        balanced sub-problem.
\end{enumerate}
The RBF kernel is:
\begin{equation}
  K(\mathbf{x}_i, \mathbf{x}_j) =
    \exp\!\bigl(-\gamma \|\mathbf{x}_i - \mathbf{x}_j\|^2\bigr)
\end{equation}
Hyperparameters $C \in \{1, 10, 100\}$ and
$\gamma \in \{\text{scale}, 0.01\}$ are selected via 3-fold
cross-validation optimising macro-F1.

\subsubsection{Model 3: Left-Right HMM}

For the HMM we use frame-level MFCC sequences
($T \times 13$ per clip, using only the first 13 MFCCs to reduce
noise). One left-right Gaussian HMM with 8 hidden states is trained
per class. The left-right topology enforces:
\begin{equation}
  a_{ij} = 0 \quad \text{for } j < i
  \qquad\text{(no backward transitions)}
\end{equation}
with self-loop probability 0.7 and forward transition probability 0.3.
This mirrors the temporal progression of a cry episode: onset
$\rightarrow$ peak $\rightarrow$ offset.

\subsubsection{Model 4: Random Forest}

500 decision trees with \texttt{class\_weight='balanced\_subsample'},
\texttt{max\_features='sqrt'}, and \texttt{min\_samples\_leaf=2}.
Feature importance scores $\text{MDI}(f) = \sum_t \Delta \text{Gini}(f, t)$
identify which acoustic features drive classification.

\subsubsection{Model 5: XGBoost}

300 boosting rounds with max depth 6, learning rate 0.1, and balanced
sample weights computed as $w_i = N / (K \cdot n_{y_i})$. Column and
row subsampling (0.8 each) provides regularisation.

\subsubsection{Model 6: Stacking Ensemble}

A meta-learner (Logistic Regression with balanced class weights)
combines the probability outputs of SVM, RF, and XGBoost:
\begin{equation}
  \mathbf{z}_i = [\hat{p}_{\text{SVM}}(\mathbf{x}_i),\;
                   \hat{p}_{\text{RF}}(\mathbf{x}_i),\;
                   \hat{p}_{\text{XGB}}(\mathbf{x}_i)] \in \mathbb{R}^{15}
\end{equation}
The meta-learner is trained on the same training set (no cross-validated
meta-features for computational efficiency in Phase~1; cross-validated
stacking will be used in Phase~3).

\subsection{Evaluation Protocol}
\label{sec:eval_protocol}

The dataset is split 80/20 (train/test) with stratification by class
and fixed random seed (42) for reproducibility. We report:
\begin{itemize}[nosep]
  \item Per-class precision, recall, and F1-score
  \item \textbf{Macro-averaged F1} (primary metric---treats all classes
        equally regardless of support)
  \item Weighted-averaged F1
  \item Overall accuracy
  \item Matthews Correlation Coefficient (MCC)
  \item Cohen's Kappa ($\kappa$)
  \item AUC-ROC (one-vs-rest, macro-averaged)
  \item Normalised confusion matrices
\end{itemize}

\paragraph{Why macro-F1 over accuracy.}
With imbalanced classes, accuracy is misleading: predicting ``hunger''
for every input yields $\sim$27\% accuracy on the augmented set
(or $\sim$84\% on the original). Macro-F1 weights each class equally
and penalises models that ignore minority classes.

% ======================================================================
\section{Results}
\label{sec:results}

\subsection{Data Leakage Analysis}
\label{sec:leakage}

Our initial pipeline augmented audio to disk \textit{before} splitting,
then applied a random 80/20 train/test split on all files (originals +
augmented). This caused augmented variants of the same original
recording to appear in both train and test sets. For example, the 8
original burping clips were augmented to 144 files; a random split
placed $\sim$115 in training and $\sim$29 in testing---but each test
sample was a pitch-shifted or time-stretched copy of a training sample.

\begin{table}[H]
\centering
\caption{Impact of data leakage: Leaky vs.\ leak-free evaluation.}
\label{tab:leakage}
\begin{tabular}{lcccc}
\toprule
\textbf{Model} & \multicolumn{2}{c}{\textbf{Leaky (Before Fix)}} & \multicolumn{2}{c}{\textbf{Leak-Free (After Fix)}} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5}
 & Acc & MacF1 & Acc & MacF1 \\
\midrule
XGBoost   & 0.965 & 0.965 & 0.804 & 0.178 \\
SVM       & 0.926 & 0.921 & 0.815 & 0.270 \\
RF        & 0.929 & 0.929 & 0.804 & 0.179 \\
Ensemble  & 0.951 & 0.944 & 0.837 & 0.249 \\
GMM       & 0.622 & 0.547 & 0.641 & 0.207 \\
HMM       & 0.348 & 0.310 & 0.674 & 0.162 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Key observation.}
Discriminative models (XGBoost, SVM, RF) suffered the largest drops
because they had the most capacity to memorise near-duplicate patterns.
The leaky XGBoost macro-F1 of 0.965 collapsed to 0.178---a
\textbf{81.6\%} decrease---confirming that the original results were
almost entirely driven by test-set contamination. Generative models
(GMM, HMM) were less affected because they model class distributions
rather than decision boundaries, making them less susceptible to
memorising augmented variants.

\subsection{Leak-Free Model Comparison}

All results below use the corrected split-before-augment protocol.
The test set consists of 92 original recordings with no augmented
variants.

\begin{table}[H]
\centering
\caption{Full evaluation metrics (leak-free, 92-sample test set).}
\label{tab:results_full}
\begin{tabular}{lcccccc}
\toprule
\textbf{Model} & \textbf{Acc} & \textbf{MacF1} & \textbf{WgtF1} & \textbf{MCC} & \textbf{$\kappa$} & \textbf{AUC} \\
\midrule
Bayesian GMM    & 0.641 & 0.207 & 0.670 & $-$0.045 & $-$0.044 & 0.454 \\
\textbf{SVM (SMOTE)} & \textbf{0.815} & \textbf{0.270} & \textbf{0.783} & \textbf{0.216} & \textbf{0.203} & \textbf{0.707} \\
HMM (left-right) & 0.674 & 0.162 & 0.678 & $-$0.080 & $-$0.079 & --- \\
Random Forest   & 0.804 & 0.179 & 0.751 & 0.013 & 0.010 & 0.632 \\
XGBoost         & 0.804 & 0.178 & 0.746 & $-$0.045 & $-$0.032 & 0.661 \\
Ensemble        & 0.837 & 0.249 & 0.780 & 0.155 & 0.096 & 0.599 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Interpretation.}
SVM achieves the highest macro-F1 (0.270), while Ensemble achieves the
highest accuracy (83.7\%). However, the near-zero MCC and $\kappa$ values
for most models indicate that performance is dominated by majority-class
(hunger) prediction. The accuracy metric is misleading: a trivial
``always predict hunger'' classifier would achieve $77/92 = 83.7\%$
accuracy.

\subsection{Per-Class Analysis}

\begin{table}[H]
\centering
\caption{Per-class F1 scores (leak-free evaluation).}
\label{tab:perclass}
\begin{tabular}{lccccc}
\toprule
\textbf{Model} & \textbf{Hunger} & \textbf{Belly Pain} & \textbf{Burping} & \textbf{Discomfort} & \textbf{Tiredness} \\
 & (77) & (3) & (2) & (5) & (5) \\
\midrule
Bayesian GMM   & 0.78 & 0.00 & 0.00 & 0.25 & 0.00 \\
SVM (SMOTE)    & 0.91 & 0.00 & 0.00 & 0.44 & 0.00 \\
HMM            & 0.81 & 0.00 & 0.00 & 0.00 & 0.00 \\
Random Forest  & 0.90 & 0.00 & 0.00 & 0.00 & 0.00 \\
XGBoost        & 0.89 & 0.00 & 0.00 & 0.00 & 0.00 \\
Ensemble       & 0.91 & 0.00 & 0.00 & 0.33 & 0.00 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Critical finding.}
All models achieve zero F1 on belly\_pain (3 test samples),
burping (2 test samples), and tiredness (5 test samples). Only SVM
recognises discomfort ($F_1 = 0.44$). This reveals the fundamental
limitation: with only 2--5 test samples per minority class, classical
ML models trained on hand-crafted features lack the representational
capacity to generalise from such limited original data, even with
training-set augmentation.

\subsection{Failure Analysis}
\label{sec:failure}

\paragraph{Majority-class collapse.}
All six models exhibit majority-class collapse: they predict
``hunger'' for nearly all inputs because hunger dominates both
training (305/1505 original + augmented) and testing (77/92 = 83.7\%).
The high accuracy ($\sim$80--84\%) masks the fact that minority classes
are completely ignored (macro-F1 $\leq 0.27$).

\paragraph{Augmentation cannot replace real data.}
Despite augmenting burping from 8 to 300 training samples,
all models achieve $F_1 = 0.00$ on the 2 original burping test samples.
The 17 augmentation transforms (pitch shift, time stretch, noise) modify
surface-level acoustics but preserve the underlying spectral template.
With only 8 original templates, the augmented training set lacks the
spectral diversity needed for generalisation.

\paragraph{GMM struggles with high dimensionality.}
The Bayesian GMM achieves only 64.1\% accuracy, with the Dirichlet
process collapsing most components to a single effective component per
class. The 411-dimensional feature space is too sparse for reliable
density estimation with limited data.

\paragraph{SVM is the only model recognising minority classes.}
SVM achieves $F_1 = 0.44$ on discomfort (the only non-zero minority
F1 across all models), likely because the SMOTE + OvO + balanced
class-weight triple strategy provides the most robust imbalance
handling. The OvO decomposition is particularly beneficial, reducing
each sub-problem to a binary decision where the minority class is
not overwhelmed.

% ======================================================================
\section{Feature Importance Analysis}
\label{sec:feature_importance}

Random Forest's MDI (Mean Decrease in Impurity) identifies the most
discriminative features:

\begin{table}[H]
\centering
\caption{Top 10 most important features (Random Forest MDI, leak-free).}
\label{tab:feature_imp}
\begin{tabular}{clc}
\toprule
\textbf{Rank} & \textbf{Feature} & \textbf{Importance} \\
\midrule
1  & spectral\_contrast\_1\_std & 0.0127 \\
2  & mfcc\_2\_std              & 0.0117 \\
3  & spectral\_contrast\_2\_mean & 0.0108 \\
4  & mfcc\_9\_mean             & 0.0103 \\
5  & voiced\_fraction          & 0.0090 \\
6  & cqcc\_4\_mean             & 0.0085 \\
7  & delta\_2\_std             & 0.0079 \\
8  & f0\_min                   & 0.0078 \\
9  & rms\_energy               & 0.0073 \\
10 & delta2\_7\_std            & 0.0073 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Key observations.}
Spectral contrast features rank highest, suggesting that harmonic
structure (distinguishing modal phonation from dysphonation) is the
most discriminative acoustic property. The MFCC variability features
(std) outrank the mean values, indicating that \textit{temporal
instability} of the spectral envelope is more informative than its
average shape. The voiced fraction ($F_0$-based) ranks 5th,
confirming the diagnostic value of prosodic features. Notably,
feature importances are more uniformly distributed compared to the
leaky evaluation, where a few features dominated---consistent with
a model that is genuinely learning rather than memorising.

% ======================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Bias--Variance Analysis}

The model hierarchy reveals a clear bias--variance tradeoff:
\begin{itemize}[nosep]
  \item \textbf{GMM}: High bias (Gaussian assumption is only partially
        satisfied), low variance---performs consistently but with a
        performance ceiling.
  \item \textbf{SVM}: Moderate bias (kernel flexibility), but
        sensitive to the $C/\gamma$ tradeoff and class imbalance.
  \item \textbf{RF/XGBoost}: Low bias (nonparametric), controlled
        variance via bagging/boosting---best individual performers.
  \item \textbf{Ensemble}: Lowest variance via model averaging,
        achieving the best generalisation.
\end{itemize}

\subsection{Theoretical Assumptions}

\paragraph{I.I.D.\ assumption and data leakage.}
Our initial pipeline violated the i.i.d.\ assumption by allowing
augmented variants of the same original to appear in both train and
test sets. The corrected split-before-augment protocol ensures
strict separation: augmentation is applied only to the training
partition, and the test set contains exclusively original recordings.
This produces a pessimistic but honest evaluation.

\paragraph{Gaussianity violations.}
The 28\% of feature--class combinations that fail the Shapiro--Wilk
test correspond primarily to spectral contrast and chroma features,
which exhibit heavy-tailed distributions. This partially explains
GMM's lower performance compared to distribution-free methods (RF,
XGBoost).

\subsection{Limitations}

\begin{enumerate}[nosep]
  \item \textbf{Extreme class imbalance}: The 47:1 ratio between
        hunger (382) and burping (8) means the leak-free test set
        contains only 2--5 samples per minority class. Per-class
        metrics have very high variance at these sample sizes.
  \item \textbf{Augmentation cannot substitute real data}: All 17
        transforms operate on the same spectral templates, limiting
        diversity. Cross-corpus data (Baby Chillanto, CryCeleb) would
        provide genuinely new acoustic patterns.
  \item \textbf{Clip-level features}: Summarising temporal dynamics
        into per-clip statistics loses information about cry
        progression patterns that sequence models (LSTM, Transformer)
        can capture.
  \item \textbf{Dataset homogeneity}: All clips derive from the
        Donate-a-Cry corpus. Real-world deployment will face domain
        shift from background noise, microphone variability, and
        infant individuality.
\end{enumerate}

\subsection{Phase~2 and Phase~3 Roadmap}

\paragraph{Phase~2 (Deep Learning).}
The leak-free results demonstrate that hand-crafted features with
classical ML are insufficient for minority-class recognition in this
data regime. Phase~2 will train CNN architectures on log-Mel
spectrograms (64 mel bands $\times$ 350 time frames), leveraging
transfer learning from pretrained audio models (e.g., MobileNetV2,
YAMNet) to overcome the limited training data. SpecAugment (time and
frequency masking) and mixup augmentation will provide additional
regularisation. Focal loss will replace cross-entropy to focus
learning on hard minority-class examples.

\paragraph{Phase~3 (Hybrid + Edge Deployment).}
Decision-level fusion of Phase~1 probabilistic scores with Phase~2
neural embeddings, deployed on ESP32 via TFLite Micro and in browsers
via ONNX.js. The hybrid architecture enables an ablation study:
Model~A (classical ML only, macro-F1 $= 0.27$) vs.\ Model~B (DL only)
vs.\ Model~C (hybrid fusion, target: outperform both).

% ======================================================================
\section{Conclusion}
\label{sec:conclusion}

This phase establishes a rigorous, leak-free classical ML baseline for
five-class infant cry classification. The key contributions are:

\begin{enumerate}[nosep]
  \item A \textbf{411-dimensional neuroscience-informed feature vector}
        combining MFCCs, CQCCs, F0 dynamics, spectral contrast, and
        chroma---grounded in the acoustic science of infant cry
        phonation modes.
  \item \textbf{Identification and correction of data leakage} in the
        augmentation pipeline, which inflated prior results from
        macro-F1 $= 0.965$ to a corrected $0.270$---an 81.6\% decrease
        that reveals the true difficulty of this task under extreme
        class imbalance.
  \item \textbf{Split-before-augment protocol} ensuring test-set
        integrity: originals split first, augmentation applied only to
        training data in memory.
  \item \textbf{Honest baseline metrics}: SVM (macro-F1 $= 0.270$,
        AUC $= 0.707$) provides the reference point for Phase~2
        deep learning improvement.
  \item \textbf{Six classical models} with comprehensive imbalance
        handling (SMOTE, OvO, balanced class weights, Bayesian priors),
        demonstrating that classical ML alone cannot overcome the
        fundamental data scarcity (8 burping originals).
\end{enumerate}

\noindent
The data leakage discovery is itself a methodological contribution:
it demonstrates the critical importance of split-before-augment
protocols in audio classification, particularly when augmentation
preserves the underlying spectral template. Phase~2 will introduce
CNN architectures on mel-spectrogram representations with transfer
learning, and Phase~3 will fuse classical and deep approaches into a
hybrid system deployable on edge devices.

% ======================================================================
\bibliographystyle{ieeetr}
\begin{thebibliography}{20}

\bibitem{dunstan2006}
P.~Dunstan,
``Dunstan Baby Language,'' 2006.
\url{https://www.dunstanbaby.com}

\bibitem{newman2007cry}
J.~D.~Newman,
``Neural circuits underlying crying and cry responding in mammals,''
\textit{Behavioural Brain Research}, vol.~182, no.~2, pp.~155--165, 2007.

\bibitem{lester1978acoustic}
B.~M.~Lester and C.~F.~Z.~Boukydis,
``Infant crying: Theoretical and research perspectives,''
\textit{Plenum Press}, New York, 1985.

\bibitem{porges2011polyvagal}
S.~W.~Porges,
\textit{The Polyvagal Theory: Neurophysiological Foundations of
Emotions, Attachment, Communication, and Self-regulation},
W.~W.~Norton, 2011.

\bibitem{ji2021review}
C.~Ji, T.~Mudiyanselage, Y.~Gao, and Y.~Pan,
``A review of infant cry analysis and classification,''
\textit{EURASIP Journal on Audio, Speech, and Music Processing},
vol.~2021, no.~1, pp.~1--17, 2021.

\bibitem{chang2017infant}
C.-Y.~Chang and J.-J.~Li,
``Application of deep learning for recognizing infant cries,''
in \textit{Proc.\ IEEE ICCE-TW}, pp.~53--54, 2017.

\bibitem{davis1980mfcc}
S.~Davis and P.~Mermelstein,
``Comparison of parametric representations for monosyllabic word
recognition,''
\textit{IEEE Trans.\ ASSP}, vol.~28, no.~4, pp.~357--366, 1980.

\bibitem{todisco2017cqcc}
M.~Todisco, H.~Delgado, and N.~Evans,
``Constant Q cepstral coefficients: A spoofing countermeasure for
automatic speaker verification,''
\textit{Computer Speech \& Language}, vol.~45, pp.~516--535, 2017.

\bibitem{maghfira2020gmm}
T.~N.~Maghfira, T.~Basaruddin, and A.~Krisnadhi,
``Infant cry classification using CNN--RNN,''
\textit{J.\ Phys.: Conf.\ Ser.}, vol.~1528, 012019, 2020.

\bibitem{reynolds2000speaker}
D.~A.~Reynolds, T.~F.~Quatieri, and R.~B.~Dunn,
``Speaker verification using adapted Gaussian mixture models,''
\textit{Digital Signal Processing}, vol.~10, pp.~19--41, 2000.

\bibitem{chang2011libsvm}
C.-C.~Chang and C.-J.~Lin,
``LIBSVM: A library for support vector machines,''
\textit{ACM TIST}, vol.~2, no.~3, pp.~1--27, 2011.

\bibitem{salehian2022infant}
M.~Salehian \textit{et al.},
``Infant cry classification using acoustic features and machine
learning,''
\textit{Applied Sciences}, vol.~12, no.~12, 6149, 2022.

\bibitem{rabiner1989hmm}
L.~R.~Rabiner,
``A tutorial on hidden Markov models and selected applications in
speech recognition,''
\textit{Proc.\ IEEE}, vol.~77, no.~2, pp.~257--286, 1989.

\bibitem{mauch2014pyin}
M.~Mauch and S.~Dixon,
``pYIN: A fundamental frequency estimator using probabilistic
threshold distributions,''
in \textit{Proc.\ IEEE ICASSP}, pp.~659--663, 2014.

\bibitem{donateacry}
Donate-a-Cry Corpus,
\url{https://github.com/gveres/donateacry-corpus}, 2015.

\bibitem{hershey2017cnn}
S.~Hershey \textit{et al.},
``CNN architectures for large-scale audio classification,''
in \textit{Proc.\ IEEE ICASSP}, pp.~131--135, 2017.

\bibitem{brazelton1995neonatal}
T.~B.~Brazelton,
\textit{Neonatal Behavioral Assessment Scale},
3rd~ed., Mac Keith Press, 1995.

\end{thebibliography}

\end{document}
